The promise of community-driven preprints in ecology and evolution

### 2024-05-02 ms/ms.qmd
Chnages text to incorperate suggestions

Publishing preprints -- papers communicating non-peer-reviewed research findings -- is now an entrenched practice across a multitude of scientific disciplines [@Ginsparg2011]. While preprint use in biology has had a slower uptake [@Berg2016], in ecology and evolution, new discipline-specific preprint servers, such as *EcoEvoRxiv*, provide a means by which ecologists and evolutionary biologists can disseminate research findings. Preprints are expected to promote open science, increase the visibility of research (e.g., [fu2019releasing]), and facilitate the rapid sharing of new discoveries that can have significant impacts on fundamental and applied knowledge globally. Preprints attempt to break down barriers to scientific publishing by: 1) increasing the speed, visibility and output of research findings; 2) helping establish precedence of research findings; 3) removing financial barriers to publication; and 4) encouraging constructive and critical feedback from the research community [@Proulx2013; @bourne2017ten; @vale2015accelerating].

@@

```{r, fig-summary}
#| label: fig-summary
#| fig-cap: Summary of articles posted to *EcoEvoRxiv*. A) Number of articles (preprints and postprints) published monthly to *EcoEvoRxiv* between 2018 and 2023. *EcoEvoRxiv* was established in June 2018 before the soft launch in November 2018. Notable milestones include when *EcoEvoRxiv* transitioned to the California Digitial Library (CDL), the acceptance of Spanish and Portuguese preprints and postprints and when the first IUCN Redlist Ecosystem report was accepted; B) Countries that are posting articles (preprints and postprints) to *EcoEvoRxiv* as inferred from the country of affiliation of the submitting author; C) Taxa used / covered in the articles posted to *EcoEvoRxiv* (*n* = 1080); D) Types of preprints accepted on *EcoEvoRxiv* (*n* = 620). E) Career stage of authors posting to *EcoEvoRxiv* along with the gender of the submitting author.
#| echo: false
#| warning: false
#| message: false

@@

  dates_preprints <- range(preprints$preprint_published_date)
##---------------------##
# Articles on EcoEvoRxiv over time
##---------------------##
  angle = 45
  p1 <- ggplot(data2, aes(x = preprint_pub_date)) +
    geom_histogram(fill = "black", color = "black") +
    labs(x = "Published Date", y = "Number of Papers") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(plot.tag = element_text(size = 16, face = "bold"), plot.tag.location = "margin") + ylim(0, 130) + 
      # Soft Launch of EcoEvoRxiv
    annotate( "segment", x = as.POSIXct("2018-07-01"),  xend = as.POSIXct("2018-07-01"), y = 0, yend = 82, color = "#ffd966", linetype = "dashed", size = 1.5) + 
    annotate("text", x = as.POSIXct("2018-07-01"), y = 85, label = "Soft Launch", color = "#ffd966", size = 4, angle = angle, hjust = 0) +
    # Launch of EcoEvoRxiv
    annotate( "segment", x = as.POSIXct("2018-11-11"),  xend = as.POSIXct("2018-11-11"), y = 0, yend = 82, color = "#2986cc", linetype = "dashed", size = 1.5) + 
    annotate("text", x = as.POSIXct("2018-11-11"), y = 85, label = "Launch", color = "#2986cc", size = 4, angle = angle, hjust = 0) +
    # Spanish and Port
    annotate( "segment", x = as.POSIXct("2023-02-14"),  xend = as.POSIXct("2023-02-14"), y = 0, yend = 82, color = "#3d85c6", linetype = "dashed", size = 1.5) + 
    annotate("text", x = as.POSIXct("2022-11-14"), y = 85, label = "Spanish and Portug.", color = "#3d85c6", size = 4, angle = angle, hjust = 0) +
    # Transition to CDL
    annotate( "segment", x = as.POSIXct("2022-09-14"),  xend = as.POSIXct("2022-09-14"), y = 0, yend = 82, color = "#008000", linetype = "dashed", size = 1.5) + 
    annotate("text", x = as.POSIXct("2022-07-14"), y = 85, label = "CDL", color = "#008000", size = 4, angle = angle, hjust = 0) +
   # IUCN Red Eco submission 29 August 2023
   # 1000 preprints
    annotate( "segment", x = as.POSIXct("2023-08-29"),  xend = as.POSIXct("2023-08-29"), y = 0, yend = 82, color = "#cc0000", linetype = "dashed", size = 1.5) + 
    annotate("text", x = as.POSIXct("2023-04-29"), y = 85, label = "IUCN Report", color = "#cc0000", size = 4, angle = angle, hjust = 0) 
##---------------------##
# Countries using EcoEvoRxiv
##---------------------##
  # Create a global map with countried highlighted and shaded according to the frequency of preprints from that country
  # First, we need to get the country names and the number of preprints from each country
    countries <- data2  %>% tabyl(submitting_author_country)  %>% rename("country" = submitting_author_country, "n" = n)  %>% mutate(country = ifelse(country == "NA", "Unknown", country))  %>% arrange(desc(n))  %>% mutate(country=str_to_title(country))

@@

After removing `r dim(dupl)[1]` duplicate titles – suggesting a few authors created multiple submissions for the same preprint rather than updating an existing submission – we had data for a total of `r pr` articles with ~55-60 preprints published each month in the last two years ([@fig-summary]A). Authors from a total of `r length(unique(countries$country))-1` countries have used *EcoEvoRxiv* as a preprint server. Unsurprisingly, authors from countries in North America and Europe upload the most preprints. Authors from countries in Africa, Central America and parts of Asia still do not use *EcoEvoRxiv* much ([@fig-summary]B). Overall, 90% of the articles come from just `r nrow(most_use)-1` countries. Articles covered all major taxonomic groups, with the most common being vertebrates (`r sum_taxa$perc[6]`%), plants (`r sum_taxa$perc[1]`%) and invertebrates (`r sum_taxa$perc[5]`%) ([@fig-summary]C).

@@

Given that *EcoEvoRxiv* is a community-driven preprint server it has managed to diversify the types of articles accepted. It accepts reports and books that are often not easily accessible, but may contain important data of relevance to research syntheses or for practitioners.  As a result, articles on *EcoEvoRxiv* are more diverse in nature than those on other preprint servers; nonetheless, typical research articles (`r sum_types[sum_types$type == "research article",]$perc`%) are still the most common type of preprint, followed by reviews and meta-analyses (`r sum_types[sum_types$type == "reviews and meta-analyses",]$perc`%) and opinion papers (`r sum_types[sum_types$type == "opinion",]$perc`%) ([@fig-summary]D).

Accepting a greater diversity of article types allows *EcoEvoRxiv* to begin dealing with the 'grey literature' problem whereby data that are relevant for research syntheses are not published in typical peer-reviewed journals[@haddaway2015shades; @haddaway2020eight]. This is reflected by `r nt`% of articles that could be considered ‘grey literature’ in ecology and evolutionary biology: books, book chapters, reports, and other research output types. Currently, *EcoEvoRxiv* does not host many reports (`r sum_types[sum_types$type == "report", "perc"]`%), particularly from government and industry. We would encourage authors to consider posting these types of articles to ensure they are openly accessible and more easily found. *EcoEvoRxiv* has already formed fruitful partnerships with the International Union for Conservation of Nature (IUCN) to be a place where their reports can be lodged to make them more easily accessible to the broader research community. For example, IUCN Red Ecosystem Reports are now posted to *EcoEvoRxiv* and our community has been able to work closely with the IUCN to ensure these documents meet the IUCN requirements.

@@

A significant barrier to the communication of research findings is due to the fact that they are primarily communicated in English [@amano2016languages; @amano2013four; @amano2021ten]. Communication of research through a single language has major consequences for the global distribution of knowledge resulting in knowledge gaps in some of the most biodiverse and threatened regions in the world [@amano2023role; @zenni2023multi]. Such gaps also have flow on effects for research syntheses and meta-analyses because they create a distorted picture of our knowledge base that can lead to bias and inadequate sampling that affects policy development and decision making [@hannah2024language; @zenni2023multi; @white2021geographical; @konno2020ignoring].

*EcoEvoRxiv* is the only preprint server breaking down language barriers to scientific communication; currently accepting English, Spanish, Portuguese and French manuscripts. Such initiatives are incredibly important if we are to begin filling global voids of scientific knowledge. Multilingual initiatives have been slow to take off on *EcoEvoRxiv*, with only a few Spanish preprints posted to date, and none in other accepted languages (e.g., French and Portuguese). Part of the challenge in getting authors to submit non-English preprints is the lack of awareness of *EcoEvoRxiv* in non-English speaking countries and the fact that it accepts preprints in different languages, cultural differences in the perception of preprints and a strong reliance on traditional publishing models, which typically favour publishing in English.

@@

Research papers take time to be published (see below). Early and mid-career researchers (ECMRs) need their research to come out quickly for job applications, promotion and obtaining grants to accelerate their careers [@vale2015accelerating; @sarabipour2019value]. Preprints are one way ECMRs can achieve faster dissemination and greater visibility. As such, we expect ECMRs to accept and adopt preprints more than colleagues at later career stages. At the same time, gender differences in preprint use and publication outcomes  have been observed in a number of other research fields, including ecology and evolution [@fox2019gender; @wehner2020comparison]. We therefore expect such discrepancies to manifest in preprint use on *EcoEvoRxiv*, but it is unclear to what extent. Understanding gender publishing patterns is challenging with observational data such as ours because we cannot know the gender of authors for certainty but we can use a data-driven approach to ascertain the probability that a particular name is of a given gender (male or female). To obtain an idea of an author's gender, we used the R package *gender* (vers. `r utils::packageVersion("gender")`; @mullen2021predict) to predict the most likely gender of the submitting author of a preprint. We only used algorithm-assigned names where the gender was identified with 95% certainty. For the remaining names, we performed manual searches to determine gender based on the pronouns and photographs from professional and personal websites (we acknowledge that our approach does not capture self-assigned and non-binary genders). Our data on gender has only two missing values – one where the first name of the submitting author was missing and the other one for a collective submission.

>>>>

Publishing preprints -- papers communicating non-peer-reviewed research findings -- is now an entrenched practice across a multitude of scientific disciplines [@Ginsparg2011]. While preprint use in biology has had a slower uptake [@Berg2016], in ecology and evolution, new discipline-specific preprint servers, such as *EcoEvoRxiv*, provide a means by which ecologists and evolutionary biologists can disseminate research findings. Preprints are expected to promote open science, increase the visibility of research (e.g., [@fu2019releasing]), and facilitate the rapid sharing of new discoveries that can have significant impacts on fundamental and applied knowledge globally. Preprints attempt to break down barriers to scientific publishing by: 1) increasing the speed, visibility and output of research findings; 2) helping establish precedence of research findings; 3) removing financial barriers to publication; and 4) encouraging constructive and critical feedback from the research community [@Proulx2013; @bourne2017ten; @vale2015accelerating].

@@

```{r, fig-summary}
#| label: fig-summary
#| fig-cap: Summary of articles posted to *EcoEvoRxiv*. A) Number of articles (preprints and postprints) published monthly to *EcoEvoRxiv* between 2018 and 2023. *EcoEvoRxiv* was established in June 2018 before the soft launch in November 2018. Notable milestones include when *EcoEvoRxiv* transitioned to the California Digital Library (CDL), the acceptance of Spanish and Portuguese preprints and postprints and when the first IUCN Red list Ecosystem report was accepted; B) Countries that are posting articles (preprints and postprints) to *EcoEvoRxiv* as inferred from the country of affiliation of the submitting author; C) Taxa used / covered in the articles posted to *EcoEvoRxiv* (*n* = 1080); D) Types of preprints accepted on *EcoEvoRxiv* (*n* = 620). E) Career stage of authors posting to *EcoEvoRxiv* along with the gender of the submitting author.
#| echo: false
#| warning: false
#| message: false

@@

  dates_preprints <- range(preprints$preprint_published_date)
##---------------------##
# Articles on EcoEvoRxiv over time
##---------------------##
  angle = 45
  p1 <- ggplot(data2, aes(x = preprint_pub_date)) +
    geom_histogram(fill = "black", color = "black") +
    labs(x = "Published Date", y = "Number of Papers") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(plot.tag = element_text(size = 16, face = "bold"), plot.tag.location = "margin") + ylim(0, 130) + 
      # Soft Launch of EcoEvoRxiv
    annotate( "segment", x = as.POSIXct("2018-06-01"),  xend = as.POSIXct("2018-07-01"), y = 0, yend = 82, color = "#ffd966", linetype = "dashed", size = 1.5) + 
    annotate("text", x = as.POSIXct("2018-06-01"), y = 85, label = "Soft Launch", color = "#ffd966", size = 4, angle = angle, hjust = 0) +
    # Launch of EcoEvoRxiv
    annotate( "segment", x = as.POSIXct("2018-11-11"),  xend = as.POSIXct("2018-11-11"), y = 0, yend = 82, color = "#2986cc", linetype = "dashed", size = 1.5) + 
    annotate("text", x = as.POSIXct("2018-11-11"), y = 85, label = "Launch", color = "#2986cc", size = 4, angle = angle, hjust = 0) +
    # Spanish and Port
    annotate( "segment", x = as.POSIXct("2023-02-14"),  xend = as.POSIXct("2023-02-14"), y = 0, yend = 82, color = "#3d85c6", linetype = "dashed", size = 1.5) + 
    annotate("text", x = as.POSIXct("2022-11-14"), y = 85, label = "Spanish and Portug.", color = "#3d85c6", size = 4, angle = angle, hjust = 0) +
    # Transition to CDL
    annotate( "segment", x = as.POSIXct("2022-09-14"),  xend = as.POSIXct("2022-09-14"), y = 0, yend = 82, color = "#008000", linetype = "dashed", size = 1.5) + 
    annotate("text", x = as.POSIXct("2022-07-14"), y = 85, label = "CDL", color = "#008000", size = 4, angle = angle, hjust = 0) +
   # IUCN Red Eco submission 29 August 2023
   # 1000 preprints
    annotate( "segment", x = as.POSIXct("2023-08-29"),  xend = as.POSIXct("2023-08-29"), y = 0, yend = 82, color = "#cc0000", linetype = "dashed", size = 1.5) + 
    annotate("text", x = as.POSIXct("2023-04-29"), y = 85, label = "IUCN Report", color = "#cc0000", size = 4, angle = angle, hjust = 0) 
##---------------------##
# Countries using EcoEvoRxiv
##---------------------##
  # Create a global map with countried highlighted and shaded according to the frequency of preprints from that country
  # First, we need to get the country names and the number of preprints from each country
    countries <- data2  %>% tabyl(submitting_author_country)  %>% rename("country" = submitting_author_country, "n" = n)  %>% mutate(country = ifelse(country == "NA", "Unknown", country))  %>% arrange(desc(n))  %>% mutate(country=str_to_title(country))

@@

After removing `r dim(dupl)[1]` duplicate titles – suggesting a few authors created multiple submissions for the same preprint rather than updating an existing submission – we had data for a total of `r pr` articles with ~55--60 preprints published each month in the last two years ([@fig-summary]A).  *EcoEvoRxiv* hosts papers from authors based in `r length(unique(countries$country))-1` countries, with 90% of the articles coming from just `r nrow(most_use)-1` countries. North America, Australia, and Europe upload the most preprints with much fewer coming from countries in Africa, Central America and parts of Asia ([@fig-summary]B).  Articles covered all major taxonomic groups, with the most common being vertebrates (`r sum_taxa$perc[6]`%), plants (`r sum_taxa$perc[1]`%) and invertebrates (`r sum_taxa$perc[5]`%) ([@fig-summary]C).

@@

Given that *EcoEvoRxiv* is a community-driven preprint server it has managed to diversify the types of articles accepted. It accepts reports and books that are often not easily accessible, but may contain important data of relevance to research syntheses or for practitioners.  As a result, articles on *EcoEvoRxiv* are more diverse than those on preprint servers that have more restrictive submission policies; nonetheless, typical research articles are still the most common type of preprint (`r sum_types[sum_types$type == "research article",]$perc`%), followed by reviews and meta-analyses (`r sum_types[sum_types$type == "reviews and meta-analyses",]$perc`%) and opinion papers (`r sum_types[sum_types$type == "opinion",]$perc`%) ([@fig-summary]D).

Accepting a greater diversity of article types allows *EcoEvoRxiv* to begin dealing with the 'grey literature' problem whereby data that are relevant for research syntheses are not published in typical peer-reviewed journals[@haddaway2015shades; @haddaway2020eight]. This is reflected by `r nt`% of articles on *EcoEvoRxiv* that could be considered ‘grey literature’ in ecology and evolutionary biology: books, book chapters, reports, and other research output types. Currently, *EcoEvoRxiv* does not host many reports (`r sum_types[sum_types$type == "report", "perc"]`%), particularly from government or industry. We would encourage authors to consider posting these types of articles to ensure they are openly accessible and more easily found. *EcoEvoRxiv* articles are given a unique DOI and indexed on Google Scholar. *EcoEvoRxiv* has already formed fruitful partnerships with the International Union for Conservation of Nature (IUCN) as a place where IUCN reports can be hosted to make them more easily accessible to the broader research community. For example, IUCN Red Ecosystem Reports are now posted to *EcoEvoRxiv* and our community has been able to work closely with the IUCN to ensure these documents meet the IUCN requirements.

@@

A significant barrier to the communication of research findings is due to the fact that they are primarily communicated in English [@amano2016languages; @amano2013four; @amano2021ten]. Communication of research through a single language has major consequences for the global distribution of knowledge resulting in knowledge gaps across some of the most biodiverse and threatened regions in the world [@amano2023role; @zenni2023multi]. Such gaps also have flow on effects for research syntheses and meta-analyses because they create a distorted picture of our knowledge base that can lead to bias and inadequate sampling that affects policy development and decision-making [@hannah2024language; @zenni2023multi; @white2021geographical; @konno2020ignoring].

*EcoEvoRxiv* is the only preprint server to date breaking down language barriers to scientific communication; currently accepting English, Spanish, Portuguese and French manuscripts. Such initiatives are incredibly important if we are to begin filling global voids of scientific knowledge. However, multilingual initiatives have been slow to take off on *EcoEvoRxiv*, with only a few Spanish preprints posted to date, and none in other accepted languages (e.g., French and Portuguese). Part of the challenge in getting authors to submit non-English preprints is the lack of awareness of *EcoEvoRxiv* in non-English speaking countries and the fact that it accepts preprints in different languages, cultural differences in the perception of preprints, and a strong reliance on traditional publishing models which typically mandate publishing in English.

@@

Research papers can take long periods of time to become published (see below). However, Early and Mid-Career Researchers (EMCRs) are under pressure to publish rapidly to be competitive in job applications, promotions, and obtaining grants to accelerate their careers [@vale2015accelerating; @sarabipour2019value]. Preprints are one way EMCRs can achieve faster dissemination and greater visibility [@fu2019releasing]. As such, we expect EMCRs to accept and adopt preprints more than colleagues at later career stages. At the same time, gender differences in preprint use and publication outcomes  have been observed in a number of other research fields, including ecology and evolutionary biology [@fox2019gender; @wehner2020comparison]. Therefore such discrepancies are expected to manifest in preprint use on *EcoEvoRxiv*, but it is unclear to what extent. Understanding gender publishing patterns is challenging with observational data such as ours because we cannot know the gender of authors for certain but we can use a data-driven approach to ascertain the probability that a particular name is of a given gender (man or women). To obtain an idea of an author's gender, we used the R package *gender* (vers. `r utils::packageVersion("gender")`; @mullen2021predict) to predict the most likely gender of the submitting author of a preprint. We only used algorithm-assigned names where the gender was identified with 95% certainty. For the remaining names, we performed manual searches to determine gender based on the pronouns and photographs from professional and personal websites (we acknowledge that our approach does not capture self-assigned and non-binary genders). Our data on gender has only two missing values – one where the first name of the submitting author was missing and the other one for a collective submission.
